---
output:
  pdf_document: default
  html_document: default
---


```{r echo=FALSE}
library(tidyverse)
library(kableExtra)
library(effectsize)
library(agricolae)
library(car)
library(PMCMRplus)
library(additivityTests)
library(lmerTest)
library(nlme)
library(emmeans)
```

# Analysis of Covariance (ANCOVA)

In various cases, researchers are interested in comparing treatments after adjusting for a numeric predictor that is associated with an experimental unit. That is often, but not necessarily, a baseline pre-treatment score on the unit. The goal is to determine whether treatment effects are present after controlling for the covariate.

When the Analysis of Covariance was first developed, the computations used were very complicated. Now, it's just as simple as fitting a regression with one or more predictors and dummy variables for the treatment conditions. Models can have one or more factors. We will consider the case with one treatment factor and one covariate.

## Additive Model 

In the additive model, the slope relating the covariate to the response is assumed to be the same for each treatment. As with the 1-Way ANOVA model, we will use $r$ as the number of treatments with $n_i$ replicates for treatment $i$. 
We will use centered $X$ values for the covariate for ease of interpretation when describing adjusted means. We will also include $r-1$ indicators for treatments $1,\ldots,r-1$ assuming again that treatment effects sum to 0.

$$
Y_{ij}= \mu_{\bullet} + \tau_1 I_{ij1} + \cdots +
\tau_{r-1}I_{ij,r-1} + 
\gamma \left(X_{ij}-\overline{X}_{\bullet\bullet}\right) +\epsilon_{ijk}
$$

$$
\epsilon_{ijk} \sim NID\left(0,\sigma^2\right) \qquad
\sum_{i=1}^r\tau_i=0 \quad \Rightarrow \quad \tau_r=-\sum_{i=1}^{r-1}\tau_i
$$
$$
I_{ij1} = 
\left\{ \begin{array}{r@{\quad:\quad}l}
    1 & i=1 \\
    0 & i=2,\ldots,r-1 \\
    -1 & i=r\\
    \end{array} \right.
  \qquad \cdots
   \qquad
I_{ij,r-1} = 
\left\{ \begin{array}{r@{\quad:\quad}l}
    1 & i=r-1 \\
    0 & i=1,\ldots,r-2 \\
    -1 & i=r\\
     \end{array} \right.
$$

To test for treatment effects, controlling for the covariate, we test
$H_0:\tau_1=\cdots = \tau_r=0$.
This is easily conducted as a general linear test, where the full model contains the $r-1$ treatment indicator variables and the centered covariate.
The reduced model contains only the centered covariate. The test is as follows.

$$
TS: F^*=\frac{\left[\frac{SSE(R)-SSE(C)}{df_R-df_C}\right]}{
\left[\frac{SSE(C)}{df_C}\right]
}
\qquad RR:F^*
\geq F_{1-\alpha;df_R-df_F,df_F}
$$
Note that for the current case, $df_R=n_T-2$ and $df_F=n_T-2-(r-1)=n_T-4$.

**Example 13.1 - Comparison of Skin Softeners**

A study compared two treatments and a control $(r=3)$ for skin softening
[@maor_1997]. The active treatments were as follows, and the covariate was the pre-treatment softness score for the subjects. There were 20 subjects per treatment, so $n_T=3(20)=60$. The response is skin roughness (lower scores are better).

- Treatment 1 - Gel Formulation
- Treatment 2 - Gel Formulation + 1% Dead Sea Concentrate
- Treatment 3 - Placebo

The R code and output are given below.

```{r ch1201, fig.cap='Analysis of Covariance for skin softener study - Additive Model', out.width='80%', fig.asp=.75, fig.align='center'}
dsm <- read.csv("data/deadseaminerals.csv")
head(dsm,3); tail(dsm,3)
dsm$I1 <- ifelse(dsm$trt_Grp==1,1,ifelse(dsm$trt_Grp==3,-1,0))
dsm$I2 <- ifelse(dsm$trt_Grp==2,1,ifelse(dsm$trt_Grp==3,-1,0))
dsm$Xc <- dsm$pre_x - mean(dsm$pre_x)

dsm.mod1 <- lm(post_y ~ I1 + I2 + Xc, data=dsm)
summary(dsm.mod1)
anova(dsm.mod1)

dsm.mod2 <- lm(post_y ~ Xc, data=dsm)
summary(dsm.mod2)
anova(dsm.mod2)

anova(dsm.mod2, dsm.mod1)

summary(dsm$Xc)

Xc.grid <- seq(-200,200,0.1)
plot(dsm$post_y ~ dsm$Xc, pch=16, col=dsm$trt_Grp, cex=0.6,
     xlab="Pre-Treatment Score", ylab="Post-Treatment Score",
     xlim=c(-200,200), ylim=c(0,320), main="Skin Roughness")
lines(Xc.grid, 
      dsm.mod1$coef[1]+dsm.mod1$coef[2]+Xc.grid*dsm.mod1$coef[4],
      col=1)
lines(Xc.grid, 
      dsm.mod1$coef[1]+dsm.mod1$coef[3]+Xc.grid*dsm.mod1$coef[4],
      col=2)
lines(Xc.grid, 
      dsm.mod1$coef[1]-dsm.mod1$coef[2]-dsm.mod1$coef[3]+Xc.grid*dsm.mod1$coef[4],
      col=3)
legend("topleft", c("Gel", "Gel+DeadSea","Placebo"), col=1:3, lty=1)
```

The $F^*$-statistic is 61.297 with 2 and 56 degrees of freedom and $P<.0001$. There is strong evidence of differences among the treatments, controlling for baseline score.

$$\nabla $$

### Adjusted means

Where the lines cross at the centered X level of 0 are the **adjusted means**. That is, they are the predicted scores for the treatments at the "average" value of the uncentered covariate. Based on this additive model,
we obtain the following adjusted means.

$$
i=1,\ldots , r-1:\quad   \hat{\mu}_i = \hat{\mu}_{\bullet} +\hat{\tau}_i \qquad 
\hat{\mu}_r=\hat{\mu}_{\bullet} -\sum_{i=1}^{r-1}\hat{\tau}_i
$$

The goal is to typically compare the adjusted means, as would be done for the 1-Way ANOVA. The standard errors depend on the variance-covariance matrix of the regression coefficients, which is obtained with the **vcov**
function from the regression model fit. The variance-covariance matrix for this example
is given below.

$$
\begin{bmatrix}
\hat{V}\left\{\hat{\mu}_{\bullet}\right\} &
\hat{\mbox{COV}}\left\{\hat{\mu}_{\bullet},\hat{\tau}_1\right\} &
\hat{\mbox{COV}}\left\{\hat{\mu}_{\bullet},\hat{\tau}_2\right\} &
\hat{\mbox{COV}}\left\{\hat{\mu}_{\bullet},\hat{\gamma}\right\} \\
\hat{\mbox{COV}}\left\{\hat{\mu}_{\bullet},\hat{\tau}_1\right\} &
\hat{V}\left\{\hat{\tau}_1\right\} &
\hat{\mbox{COV}}\left\{\hat{\tau}_1,\hat{\tau}_2\right\} &
\hat{\mbox{COV}}\left\{\hat{\tau}_1,\hat{\gamma}\right\} \\
\hat{\mbox{COV}}\left\{\hat{\mu}_{\bullet},\hat{\tau}_2\right\} &
\hat{\mbox{COV}}\left\{\hat{\tau}_1,\hat{\tau}_2\right\} &
\hat{V}\left\{\hat{\tau}_2\right\} &
\hat{\mbox{COV}}\left\{\hat{\tau}_2,\hat{\gamma}\right\} \\
\hat{\mbox{COV}}\left\{\hat{\mu}_{\bullet},\hat{\gamma}\right\} &
\hat{\mbox{COV}}\left\{\hat{\tau}_1,\hat{\gamma}\right\} &
\hat{\mbox{COV}}\left\{\hat{\tau}_2,\hat{\gamma}\right\} &
\hat{V}\left\{\hat{\gamma}\right\}
\end{bmatrix}
$$

$$
i,i' < r  \quad
\hat{SE}\left\{\hat{\tau}_i-\hat{\tau}_{i'}\right\} =
\sqrt{\hat{V}\left\{\hat{\tau}_i\right\}+\hat{V}\left\{\hat{\tau}_{i'}\right\}
-2\hat{\mbox{COV}}\left\{\hat{\tau}_i,\hat{\tau}_{i'}\right\}}
$$

When comparing the first $r-1$ treatments with the $r^{th}$, we use the following result.

$$
\hat{\tau}_i-\hat{\tau}_r=\hat{\tau}_i-\left(-\sum_{i=1}^{r-1}\hat{\tau}_i\right)=
2\hat{\tau}_i + \sum_{\stackrel{i'=1}{i'\ne i}}^{r-1}\hat{\tau}_{i'}
$$

Then the (messy) standard errors for these differences are as follow.

$$
\hat{SE}\left\{\hat{\tau}_i-\hat{\tau}_r\right\} =
\sqrt{4\hat{V}\left\{\hat{\tau}_i\right\} +
\sum_{\stackrel{i'=1}{i'\ne i}}^{r-1}\hat{V}\left\{\hat{\tau}_{i'}\right\}
+4\sum_{\stackrel{i'=1}{i'\ne i}}^{r-1}\hat{\mbox{COV}}\left\{\hat{\tau}_i,\hat{\tau}_{i'}\right\}
+2\sum_{\stackrel{i'=1}{i'\ne i}}^{r-2}\sum_{\stackrel{i''=i'+1}{i''\ne i}}^{r-1}
\hat{\mbox{COV}}\left\{\hat{\tau}_{i'},\hat{\tau}_{i''}\right\}
}
$$
Here, we consider a few cases to observe these variance calculations.
When $r=2$, we have the following simple case.

$$
\hat{\tau}_2 = -\hat{\tau}_1 \quad \Rightarrow \quad 
\hat{\tau}_1-\hat{\tau}_2=
2\hat{\tau}_1 
\quad \Rightarrow \quad 
\hat{V}\left\{\hat{\tau}_1-\hat{\tau}_2\right\}
=4\hat{V}\left\{\hat{\tau}_1\right\}
$$
When $r=3$, as in the skin softener study, we have the following results.

$$
\hat{\tau}_3 = -\hat{\tau}_1-\hat{\tau}_2 \quad \Rightarrow \quad
\hat{\tau}_1-\hat{\tau}_3 = \hat{\tau}_1-\left[\hat{\tau}_1-\hat{\tau}_2\right]=
2\hat{\tau}_1+\hat{\tau}_2 
$$
$$\quad \Rightarrow \quad
\hat{V}\left\{\hat{\tau}_1-\hat{\tau}_3\right\}=
4\hat{V}\left\{\hat{\tau}_1\right\}+\hat{V}\left\{\hat{\tau}_2\right\}+
4\hat{\mbox{COV}}\left\{\hat{\tau}_1,\hat{\tau}_2\right\}
$$
Finally, when $r=4$, have the following results.
$$
\hat{\tau}_1-\hat{\tau}_4 = 
\hat{\tau}_1-\left[\hat{\tau}_1-\hat{\tau}_2-\hat{\tau}_3\right]=
2\hat{\tau}_1+\hat{\tau}_2+\hat{\tau}_3 \quad \Rightarrow \quad  
$$
$$
\hat{V}\left\{\hat{\tau}_1-\hat{\tau}_4\right\}=
4\hat{V}\left\{\hat{\tau}_1\right\}+\hat{V}\left\{\hat{\tau}_2\right\}+
\hat{V}\left\{\hat{\tau}_3\right\}+
4\hat{\mbox{COV}}\left\{\hat{\tau}_1,\hat{\tau}_2\right\}+
4\hat{\mbox{COV}}\left\{\hat{\tau}_1,\hat{\tau}_3\right\}+
2\hat{\mbox{COV}}\left\{\hat{\tau}_2,\hat{\tau}_3\right\}
$$


**Example 13.2 - Comparison of Skin Softeners**

We obtain the adjusted means and their differences along with multiple comparisons with the following R code.

```{r ch1202}
dsm <- read.csv("data/deadseaminerals.csv")
head(dsm,2); tail(dsm,2)
dsm$I1 <- ifelse(dsm$trt_Grp==1,1,ifelse(dsm$trt_Grp==3,-1,0))
dsm$I2 <- ifelse(dsm$trt_Grp==2,1,ifelse(dsm$trt_Grp==3,-1,0))
dsm$Xc <- dsm$pre_x - mean(dsm$pre_x)

dsm.mod1 <- lm(post_y ~ I1 + I2 + Xc, data=dsm)
summary(dsm.mod1)
anova(dsm.mod1)
vcov1 <- vcov(dsm.mod1)
r <- length(unique(dsm$trt_Grp))

adjmean <- rep(0,r)
for (i1 in 1:(r-1)) adjmean[i1] <- dsm.mod1$coef[1] + dsm.mod1$coef[i1+1]
adjmean[r] <- dsm.mod1$coef[1]-sum(dsm.mod1$coef[2:r])
adjmean

se.diff <- rep(0,r*(r-1)/2)
se.diff[1] <- sqrt(vcov1[2,2]+vcov1[3,3]-2*vcov1[2,3])
se.diff[2] <- sqrt(4*vcov1[2,2]+vcov1[3,3]+4*vcov1[2,3])
se.diff[3] <- sqrt(4*vcov1[3,3]+vcov1[2,2]+4*vcov1[2,3])

HSD.q <- qtukey(.95,r,df.residual(dsm.mod1)) 
HSD.out <- matrix(rep(0,10*r*(r-1)/2), ncol=10)
row.count <- 0
for (i1 in 2:r) {
  for (i2 in 1:(i1-1)) {
    row.count <- row.count + 1
    HSD.out[row.count,1] <- i1
    HSD.out[row.count,2] <- i2
    HSD.out[row.count,3] <- adjmean[i1]
    HSD.out[row.count,4] <- adjmean[i2]
    HSD.out[row.count,5] <- adjmean[i1] - adjmean[i2]
    HSD.out[row.count,6] <- se.diff[row.count]
    HSD.out[row.count,7] <- HSD.q*se.diff[row.count]
    HSD.out[row.count,8] <- 
      (adjmean[i1] - adjmean[i2]) - HSD.q*se.diff[row.count]
    HSD.out[row.count,9] <- 
      (adjmean[i1] - adjmean[i2]) + HSD.q*se.diff[row.count]
    HSD.out[row.count,10] <- 
      1-ptukey(sqrt(2)*abs(adjmean[i1] - adjmean[i2])/se.diff[row.count],
               r,df.residual(dsm.mod1))
  }
}

colnames(HSD.out) <- c("i", "i'", "AdjMn_i", "AdjMn_i'", "Diff", "Std Err", 
                      "HSD", "LB", "UB", "p adj")
round(HSD.out,4)
```
All three pairs of treatments are significantly different.
Treatment 2 (Gel+DeadSea) has significantly lower roughness scores than the other treatments and Gel has significantly lower scores than Placebo.

Here we use the factor variable for treatment and the **emmeans** package to obtain
and compare the adjusted means.


```{r ch1203}
dsm <- read.csv("data/deadseaminerals.csv")
head(dsm,2)
dsm$trt_Grp.f <- factor(dsm$trt_Grp, levels=1:3,labels=c("gel","gelDS", "control"))
options(contrasts=c("contr.sum", "contr.poly"))
dsm.mod.add <- aov(post_y ~ pre_x + trt_Grp.f, data=dsm)
summary.lm(dsm.mod.add)
anova(dsm.mod.add)

dsm.emm1 <- emmeans(dsm.mod.add, ~ trt_Grp.f)
dsm.emm1
contrast(dsm.emm1, method="pairwise", adjust="tukey")

```
$$\nabla  $$

## Interaction Model

When the slope relating the response to the covariate differs among treatments, the model contains interactions. This can be tested by including cross-product terms for the treatment indicator variables and the centered covariate. Then the full model containing these extra terms can be compared with the additive (reduced) model.
The interaction model is given below.

$$
Y_{ij}= \mu_{\bullet} + \sum_{k=1}^{r-1}\tau_k I_{ijk}  + 
\gamma \left(X_{ij}-\overline{X}_{\bullet\bullet}\right) +
\sum_{k=1}^{r-1}\beta_kI_{ijk}\left(X_{ij}-\overline{X}_{\bullet\bullet}\right) +
\epsilon_{ijk}
$$

Comparing adjusted means is more difficult when there is an interaction, as the difference among treatment means differs at different $X$ levels (unlike the additive model). A method was developed by Johnson and Neyman to find regions of $X$ values where two treatments are significantly different and where they are not. We will not pursue that method here, but will run the general linear test for interactions.


**Example 13.3 - Comparison of Skin Softeners**

We obtain the adjusted means and their differences along with multiple comparisons with the following R code.

```{r ch1204, fig.cap='Analysis of Covariance for skin softener study - Interaction Model', out.width='80%', fig.asp=.75, fig.align='center', echo=FALSE}
dsm <- read.csv("data/deadseaminerals.csv")
head(dsm,2); tail(dsm,2)
dsm$I1 <- ifelse(dsm$trt_Grp==1,1,ifelse(dsm$trt_Grp==3,-1,0))
dsm$I2 <- ifelse(dsm$trt_Grp==2,1,ifelse(dsm$trt_Grp==3,-1,0))
dsm$Xc <- dsm$pre_x - mean(dsm$pre_x)

dsm.mod1 <- lm(post_y ~ I1 + I2 + Xc, data=dsm)
summary(dsm.mod1)
anova(dsm.mod1)

dsm.mod2 <- lm(post_y ~ I1 + I2 + Xc + I(I1*Xc) + I(I2*Xc), data=dsm)
summary(dsm.mod2)
anova(dsm.mod2)

anova(dsm.mod1, dsm.mod2)

Xc.grid <- seq(-200,200,0.1)
plot(dsm$post_y ~ dsm$Xc, pch=16, col=dsm$trt_Grp, cex=0.6,
     xlab="Pre-Treatment Score", ylab="Post-Treatment Score",
     xlim=c(-200,200), ylim=c(0,320), main="Skin Roughness")
lines(Xc.grid, 
      dsm.mod2$coef[1]+dsm.mod2$coef[2]+Xc.grid*dsm.mod2$coef[4]
      +dsm.mod2$coef[5]*Xc.grid,
      col=1)
lines(Xc.grid, 
      dsm.mod2$coef[1]+dsm.mod2$coef[3]+Xc.grid*dsm.mod2$coef[4]
      +dsm.mod2$coef[6]*Xc.grid,
      col=2)
lines(Xc.grid, 
      dsm.mod2$coef[1]-dsm.mod2$coef[2]-dsm.mod2$coef[3]+
        Xc.grid*dsm.mod2$coef[4]-dsm.mod2$coef[5]*Xc.grid-
        dsm.mod2$coef[6]*Xc.grid,
      col=3)
legend("topleft", c("Gel", "Gel+DeadSea","Placebo"), col=1:3, lty=1)
```

The interaction is significant $(F^*=18.42, P<.0001)$. Based on the plot, there appears to be larger treatment differences among patients with higher
baseline skin roughness.

$$\nabla $$